---
title: "Publishing your analyses"
subtitle: "Persistant project repositories with the OSF"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2024-06-24
format: 
  html:
    output-file: osf.html
    number-sections: true
    toc: true
    code-overflow: wrap
    code-tools: true
    self-contained: true
  pdf:
    output-file: osf.pdf
    toc: true
    number-sections: true
    colorlinks: true
    code-overflow: wrap
  revealjs:
    output-file: osf_slides.html
    include-in-header: ../../mathjax.html # for multiple equation hyperrefs
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    # chalkboard:
    #   src: chalkboard.json
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-size: 15%
      data-background-position: 50% 92%
    fig-align: center
    fig-dpi: 300
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: ../../apa.csl
execute:
  echo: true
  eval: true
---

```{r}
#| eval: false
#| echo: false
# should be run manually
rbbt::bbt_update_bib(here::here("slides/11-osf/osf.qmd"))
```

# Learning objectives {.unlisted .unnumbered}

Today we will...

- learn about the OSF
- see popular uses of the OSF

# Resources {.unnumbered .unlisted}

# Open and FAIR data

::: {.content-visible when-format="revealjs"}

:::: columns

::: {.column width="60%"}

- recall the FAIR principles we discussed at the beginning of term
   + data should be Findable, Accessible, Interoperable, and Reusable
   + we're extending these principles to our *analyses* as well

:::

::: {.column width="40%"}
```{r}
#| echo: false
#| out-width: "95%"
#| fig-align: center
#| label: fig-FAIR-slides
#| fig-cap: "Source: [National Library of Medicine](https://www.nlm.nih.gov/oet/ed/cde/tutorial/02-300.html) (all rights reserved)"
magick::image_negate(magick::image_read(here::here("media/FAIR_nlm_nih.png")))
```
:::
::::

:::

- recall the FAIR principles we discussed at the beginning of term
   + data should be Findable, Accessible, Interoperable, and Reusable
   + we're extending these principles to our *analyses* as well
- so far, our data and analyses are stored locally on our machines
  + we need to share them with a persistant public storage
  + e.g., GitHub or GitLab, the Open Science Framework (OSF) or Zenodo

```{r}
#| echo: false
#| out-width: "50%"
#| fig-align: center
#| label: fig-FAIR
#| fig-cap: "Source: [National Library of Medicine](https://www.nlm.nih.gov/oet/ed/cde/tutorial/02-300.html) (all rights reserved)"
magick::image_read(here::here("media/FAIR_nlm_nih.png"))
```

# OSF: Open Science Framework

- we'll use the OSF ([https://osf.io/](https://osf.io/)), which is a user-friendly project management platform
  + provides persistant URLs
  + user-friendly (drag-and-drop)
  + popular for open storage of data, materials, and analyses
- also offers pre-registration and pre-print storage
- can also be connected to Dropbox, Google Drive, GitHub and GitLab
  + but this requires you to have your data and analyses stored on these services, the security of which cannot be guaranteed long-term

- if you don't already have an OSF account, click the 'sign up' button at the top right of the [OSF homepage](https://osf.io/)

## Setting up a project

- we'll start by creating a new OSF project

1. Sign in to the OSF
2. Click on 'Create new project'
    + provide a name such as 'Open Science Practices: Reproducibility coursework (SoSe2024)'
    + **Important**: set storage location to `Germany - Frankfurt`
    + add some concise description
3. Navigate to your project and explore the page and tabs

## Private or public

- you should notice near the top right corner a button 'Make Public'
  + this tells you that your project is currently private
  + this means nobody can see it but you (or any collaborators you add)
- typically you would make a repository public when it has been accepted for publication, or if you publish a pre-print
  + you can also make it public before this, but this is something to discuss with your collaborators
  
## Contributors

- contributors are typically co-authors or collaborators for a project
- click on the 'Contributors' tab (top right of the screen)
  + click on '+ Add', find my account, and add me as a collaborator with 'Read' rights
  + make sure the 'Bibliographic Contributor' button is checked
- go back to the project page, do you see any changes?

# Adding files

- our purpose for creating an OSF project was to share our data and materials
  + to do this, we navigate to the 'Files' tab
- rather unfortunately, we can only upload files (i.e., not entire folders)
  + this has the benefit of meaning our folder structure must be *intentional*
  + but the drawback that it's quite tedious if you want to share a large project

- let's start but adding our data and scripts
  + add a folder called `data`
  + and another folder called `scripts` or `code`, or whatever you prefer

## Adding data

- under `data`, add the three `csv` files you presumably have in the same folder in your project (drag and drop them, or select the big green `+`)
  + `data_lifetime_pilot`
  + `tidy_data_lifetime_pilot.csv`
  + `tidy_data_lifetime_pilot_dictionary.csv`


## Adding scripts

- under the `scripts` folder add the scripts where we worked with the data:
  + our wrangling script
  + our tables script
  + our data visualisation script
  
### Adding PDF output

- optionally, you can also upload the PDF output produced by each script
  + this means that anybody viewing your project doesn't have to download the Quarto scripts to see what was done (as they aren't viewable in-browser)

# Reproducibility

- great, so we've shared the code, not just the data!
  + this has been strongly encouraged in the reproducibility research as of late (e.g., @@laurinavichyute_share_2017)
  + but is this sufficient to ensure long-term reproducibility?
- in addition, would the code and data in their current state run on another machine?

- our focus is on sharing data and analyses with the aim of reproducibility
  + so we have to share what is necessary to make our project reproducible
- so what should we share?

## R project

- consider, for example, how we accessed the data from our scripts
  + did we use `setwd()`?
  + did we use filepaths?
  + no, we used the `here()` package within an R project
- so, we should, at minimum, also include the `.Rproj` file at the project root directory

### Reproducibility spectrum

- reproducibility is on a continuum, referred to as the *reproducibility spectrum* in @peng_reproducible_2011 (@fig-peng_2011)
  + *linked* means "*all data, metadata, and code [is] stored and linked with each other and with corresponding publications*" [@peng_reproducible_2011, p. 1227]
  + *executable* is not explained, and is more difficult to guarantee long-term as it depends on software versions

```{r}
#| echo: false
#| out-width: "100%"
#| fig-align: center
#| label: fig-peng_2011
#| fig-cap: "Source: @peng_reproducible_2011"
magick::image_read(here::here("media/peng_2011_reproducibility_spectrum.png"))
```

### Reproducibility iceberg

- a more detailed description of this continuum is givin in @rodrigues_building_nodate's *reproducibility iceberg* (@fig-rodrigues_2023)
  + our project is currently somewhere in the middle of the iceberg (or will be once it's on the OSF)
  + we're not using GitHub, with is a developer platform and useful for version control and is beyond the scope of this course
- importantly, the iceberg explicitly mentioned the `renv` package
  + this reminds us that we should be *sharing* 

```{r}
#| echo: false
#| out-width: "75%"
#| fig-align: center
#| label: fig-rodrigues_2023
#| fig-cap: "Source: @rodrigues_building_nodate"
magick::image_read(here::here("media/rodrigues_2023_reproducibility_iceberg.png"))
```

# Code review

- a code review refers to when somebody else checks your code
  + this should also include a check for reproducibility
- why should we do it?
  + firstly, everybody makes mistakes! increases the chances they'll be fixed
  + tests reproducibility

# Session Info {.unnumbered}

```{r}
print(sessionInfo(), locale = F)
```


# References {.unlisted .unnumbered visibility="uncounted"}

---
nocite: |
  @rodrigues_building_nodate
---

::: {#refs custom-style="Bibliography"}
:::