---
title: "Data communication with tables"
subtitle: "Data dictionaries and summary tables"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
footer: "Data Visualisation"
lang: en
date: 2023-04-13
format:
  revealjs: 
    output-file: slides-data_viz.html
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    # smaller: true
    scrollable: true
    slide-number: c/t
    code-link: true
    code-overflow: wrap
    code-tools: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc-title: 'Topics'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    chalkboard: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
  html:
    self-contained: true
    output-file: sheet-data_viz.html
    number-sections: true
    toc: true
    code-overflow: wrap
    code-tools: true
  pdf:
    output-file: pdf-data_viz.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
editor_options: 
  chunk_output_type: console
bibliography: ../../references.bib
csl: ../../apa.csl
---

```{r}
#| echo: false
knitr::opts_chunk$set(eval = T, # evaluate chunks
                      echo = T, # 'print code chunk?'
                      message = F, # 'print messages (e.g., warnings)?'
                      error = F, # stop when error encountered
                      warning = F) # don't print warnings
```

# Learning objectives

- create a data dictionary
- produce formatted tables with the `knitr` package
- produce summary tables

# Load packages and data

```{r}
# load tidyverse
pacman::p_load(tidyverse, here)

# load data
df_lifetime <- readr::read_csv(here::here("data/tidy_data_lifetime_pilot.csv"), 
                               # for special characters
                               locale = readr::locale(encoding = "latin1") 
                               ) |>
  mutate_if(is.character,as.factor) |> # all character variables as factor
  filter(type == "critical", # only critical trials
         px != "px3") # this participant had lots of 0's for some reason
```

# Data dictionary

- we haven't really discussed what exactly our data *are*, though
- data dictionaries (a.k.a. code books)
  + describe each variable in a dataset
  + ideally also provide information regarding possible values

## Variable names

- we can list the names of all variables in a dataset using `names()`

```{r}
#| output-location: fragment
names(df_lifetime)
```

- but we need to be able to put these names into a single column
  + where each row contains one variable name
  + and other columns contain information like description and data class

### Names to rows

```{r}
# From day 2 of Lisa DeBruine's [Coding Club: Creating an R Package](https://psyteachr.github.io/intro-r-pkgs/02-data.html#documentation)

# create as many empty strings as we name variable names
coldesc <- rep("", ncol(df_lifetime))
# add variable names to these empty strings
names(coldesc) <- names(df_lifetime)
```

```{r}
# print as code needed to create an object
dput(coldesc)
```

##

- copy the output of `dput(coldesc)` and assign it to an object
  + tip: you can reformat the code by highlighting it and using `Cmd/Ctrl+Shift+A`
    + or in the menu bar: Code > Reformat Code
- replace `c()` with `tibble()` to create a dataframe
  + and fill in the quotations with description of the data

```{r}
dict_lifetime <- tibble(
  px = "participant ID (factor)",
  trial = "trial number (ordered factor)",
  region = "sentence region (order factor)",
  region_n = "numerical representation of sentence region (ordered factor)",
  region_text = "text presented in the region (string)",
  eye = "which eye was tracking: right or left (binomial)",
  ff = "first-fixation times in milliseconds (continuous, values can be 0<)",
  fp = "first-pass reading times in milliseconds (numeric, values can be 0<)",
  rpd = "regression-path duration in milliseconds (numeric, values can be 0<)",
  tt = "total reading time in milliseconds (numeric, values can be 0<)",
  fix_count = "number of total fixations in the region (count)",
  reg_in = "whether of a regression was made into the regions (binomial: 0 = no, 1 = yes)",
  reg_in_count = "number of fixations into the region (count)",
  reg_out = "whether of a regression was made out of the regions (binomial: 0 = no, 1 = yes)",
  reg_out_count = "number of fixations out of the region (count)",
  rt = "reaction time from critical sentence presentation to button press (continuous, values can be 0<)",
  bio = "lifetime biography context sentence (string)",
  critical = "critical sentence (string)",
  gender = "gender of stimulus subject (binomial: male, female)",
  item_id = "item identification number (critical items: 1-80)",
  list = "experimental list version: base list version (1-4) and whether the yes-button was coded as 4 or 5 (factor: 14, 15, 24, 25, 34, 35, 44, 45)",
  match = "whether the referent-lifetime was congruent with tense",
  condition = "condition: lifetime (dead, alive) + tense (PP, SF) (factor)",
  name = "name of stimulis subject (string)",
  lifetime = "lifetime status of stimulus subject at time of experiment (binomial: dead, alive)",
  tense = "tense used in critical sentence (binomail: PP = present perfect, SF = simple future)",
  type = "sentence type (factor with one level: critical)",
  yes_press = "corresponding coding for the yes-button on Cedrus response box (4 = left button, 5 = right button)",
  KeyPress = "key that was pressed (4 = left button, 5 = right button)",
  accept = "whether the item was accepted, i.e., whether KeyPress equalled yes_press",
  accuracy = "whether the acceptance was accurate (reject for a mismatch, accept for a match)",
  px_accuracy = "participant's overall accuracy score"
)
```

##

- but `dict_lifetime` doesn't have the shape we want
  + each variable name is a column name
  + and its description is in the first row
  
```{r}
dict_lifetime
```

- we want to transpose the data
  + i.e., rotates the data so that the column names are in a row, with the descriptions in another row
  
## `pivot_longer()`

- takes wide data and makes it longer
  + converts headers of columns into values of a new column
  + combines the values of those columns into a new condensed column
- takes a few arguments:
  + `cols`: which columns do we want to combine into a single column?
  + `names_to`: what should we call the new column containing the previous column names?
  + `values_to`: what should we call the new column containing the values from the previous columns?

## {.unlisted .unnumbered}


::: {.content-hidden when-format="pdf"}
```{r echo = F, fig.align = "center", out.width="100%"}
#| fig-cap-location: bottom
magick::image_read(here::here("media/pivot_friends.jpeg"))
```
:::

::: {.content-visible when-format="pdf"}
```{r echo = F, fig.align = "center", out.width="50%"}
#| fig-cap-location: bottom
magick::image_read(here::here("media/pivot_friends.jpeg"))
```
:::

### Pivot our data dictionary

- this looks much better!

```{r}
#| output-location: fragment
dict_lifetime <-
  dict_lifetime |> 
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "description"
  )
```

## Save data dictionary

- now we can save our data dictionary just like we would any csv

```{r}
write_csv(dict_lifetime, here("data", "tidy_data_lifetime_pilot_dictionary.csv"))
```

# Formatted tables

- when we render our document, `dict_lifetime` won't look very pretty
- there are several packages that produce nicely formatted tables
  + `knitr`
  
```{r}
dict_lifetime |> 
  knitr::kable() 
```

## Tables as LaTeX code

- you can add the argument `"latex"` to print LaTeX code for a table in the Console
  + you can then cut and paste this code into a LaTeX (or Overleaf) script

```{r}
#| eval: false
dict_lifetime |> 
  knitr::kable("latex") 
```

- but be careful, if you're rendering to HTML the table won't be printed if you use `"latex"`

### Exercise

1. install the `knitr` package (`install.packages("knitr")`)
2. print `dict_lifetime`, but only for the following variables:
    + `px`, `trial`, `region_text`, `ff`, `fp`, and `condition`
3. use `kable()` from `knitr` to print the table

```{r}
#| echo: false
dict_lifetime |> 
  filter(variable %in% c("px", "trial", "region_text", "ff", "fp", "condition")) |> 
  knitr::kable()
```

# Data summaries

- we can create summary tables of our data

```{r, eval = F, echo = F}
# with Rmisc::summarySEwithin
df_lifetime |> 
  filter(region == "verb",
         !is.na(ff)) |> 
  Rmisc::summarySEwithin(measurevar="ff", withinvars ="condition")
```

```{r}
#| code-fold: true
# compute summary 
summary_ff <- df_lifetime |> 
  filter(region=="verb") |> 
  group_by(condition,lifetime,tense) %>%
  summarise(N = n(),
            mean.ff = mean(ff, na.rm = T),
            sd = sd(ff, na.rm = T)) %>%
  # compute standard error, confidence intervals, and lower/upper ci bounds
  mutate(se = sd / sqrt(N),
         ci = qt(1 - (0.05 / 2), N - 1) * se,
         lower.ci = mean.ff - qt(1 - (0.05 / 2), N - 1) * se,
         upper.ci = mean.ff + qt(1 - (0.05 / 2), N - 1) * se)
```

- and print the output with the `kable()` function from the `knitr` package
  + for extra customisation you can also use the `kableExtra` package (e.g., with the `kable_styling()` function)

```{r}
#| output-location: fragment

# install.packages("knitr") # if not yet installed
knitr::kable(summary_ff, digits=1,
             caption = "Table with summmary statistics for first-fixation duration at the verb region")
```

# Saving summary tables

- we could also save this table using `write_csv()`
  + but it's relatively simple to re-produce, so I wouldn't bother
  + instead, when writing up my results I would load in the data and print the summary directly
- sometimes summary tables are more code-intensive
  + in this case I would save the summary as a csv, and simply load and print it when writing in R markdown or Quarto

##

# Additional packages

There are many other packages for including tables that are publication-ready. Some that I would suggest you look into:

- [`kableExtra`](https://haozhu233.github.io/kableExtra/) which includes additionally formatting options for `knitr::kable()` tables via the `kable_styling()` function and others
  + tables must first pass through `knitr::kable()`, e.g., `my_table |> knitr::kable() |> kableExtra::kable_styling()`
- [`flextable`](https://cran.r-project.org/web/packages/flextable/index.html)
  + very flexible package for creating publication-ready tables of many formats
- `papaja::apa_table()`: the `papaja` package aids in creating APA-formatted journal articles
  + the `apa_table()` function can take objects containing results from a statisical test/model and output a formatted table
  + we'll discuss this topic more when we get into regression
  
## `kableExtra::kable_styling()`

- we'll first create a little summary table using the `iris` dataset which comes built-in with R

```{r}
sum_iris <- iris |> 
  summarise(mean = mean(Sepal.Length),
            sd = sd(Sepal.Length),
            n = n(),
            .by = Species)
```

- and we'll print the summary using `kable_styling()` (@tbl-kableExtra)

```{r}
#| label: tbl-kableExtra
#| tbl-cap: "Example output of an table using the `kableExtra` package"
sum_iris |> 
  knitr::kable() |> 
  kableExtra::kable_styling()
```

## `flextable`

- print the same summary using `flextable()` (@tbl-flextable)

```{r}
#| label: tbl-flextable
#| tbl-cap: "Example output of an table using the `flextable` package"
sum_iris |> 
  flextable::flextable()
```


## `papaja::apa_table()`

- now run a linear mixed model on the `iris` data
  
```{r}
lmm_iris <-
  lme4::lmer(Sepal.Width ~ Sepal.Length + Petal.Width +
               (1|Species), data = iris)
```

- and print a model summary table using `apa_table()` (@tbl-lmm_iris)

```{r}
#| label: tbl-lmm_iris
#| tbl-cap: "Example output of an LMM using `papaja` package"
lmm_iris |> papaja::apa_print() |> 
papaja::apa_table(caption = NULL)
```

  


# Exercise

1. create an object with some summary statistics of the variable `rt`
    + call it `summary_rt`
3. use `kable()` from `knitr` to print a table, it should look something like @tbl-rt
4. try creating the same table with one (or more) of the additional packages we saw above (`kableExtra`, `flextable`, `papaja`)

```{r}
#| echo: false
summary_rt <- df_lifetime |> 
  distinct(px,trial,.keep_all=T) |> 
  group_by(lifetime,tense,condition) %>%
  summarise(N = n(),
            mean.rt = mean(rt, na.rm = T),
            sd = sd(rt, na.rm = T))
```

```{r}
#| echo: false
#| label: tbl-rt
knitr::kable(summary_rt, digits=1,
             caption = "Summary of reaction times (ms) per condition")
```




