---
title: "The Replication Crisis"
subtitle: "And what to do about it"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2024-04-23
format: 
  html:
    output-file: replication-crisis.html
    number-sections: false
    toc: true
    code-overflow: wrap
    code-tools: true
    self-contained: true
  pdf:
    output-file: replication-crisis.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    output-file: replication-crisis_slides.html
    include-in-header: ../../mathjax.html # for multiple equation hyperrefs
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    fig-align: center
    fig-dpi: 300
editor_options: 
  chunk_output_type: console
bibliography: ../../references.bib
csl: ../../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = T, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```




# Learning Objectives {.unnumbered .unlisted}

Today we will learn about...

- the replication crisis
- replication in language sciences
- requirements for replication

# Resources {.unnumbered .unlisted}

- this lecture covers @sonning_replication_2021

- introduction article of a special issue of the Journal *Linguistics*
  + *The replication crisis: Impications for linguistics*
- contains several articles on the topic, some of which we'll read later

# Replication crisis

- data-based claims turned out to be less reliable than previously believed
  + statistical claims could not be replicated with new data
- large-scale replications brought attention to the issue
  + e.g., @nieuwland_large-scale_2018; @open_science_collaboration_estimating_2015
- this has also led to distrust in findings in academia and the public

## Most Published Research Findings are False {.smaller}

:::: {.columns}

::: {.column width="50%"}

- the issue became more widespread with @ioannidis_why_2005
  + defined bias in terms of design, analysis, and presentation factors
  + focussed on issues with *p*-values and statistical power

- @open_science_collaboration_estimating_2015 ran replications of 100 studies
  + 36% of replications found significant effects
  + 47% of original effects fell within 95% CIs of replication effect
- in essence: fewer significant findings and smaller effects in replications

- how is can this be?

:::

::: {.column width="50%"}

```{r}
#| echo: false
#| out-width: "95%"
#| fig-align: left
#| label: fig-replication
#| fig-cap: "Source: @open_science_collaboration_estimating_2015 (all rights reserved)"
magick::image_negate(magick::image_read(here::here("media/Open_Sci_collab_2015_replication-effects.png")))
```

:::

::::

## The problem with p-values

- issues relating to reported findings:
  - misuse of misinterpretation of p-values in Null Hypothesis Significant Testing [NHST\; e.g., @ioannidis_why_2005]
  - study design
  - improper use of statistical methods
    + stemming from inadequate teaching
  - selective reporting
- in other words, HARKing and p-hacking (whether consciously done or not)

## Solving the problem

- these could be mitigated with Open Science practices
  + *transparency* in writing, analyses, planning/hypothesising stages
  + *reproducibility* of analyses
  + greater value given to *replication* studies
  + embracing and addressing *uncertainty* [@vasishth_how_2021]
- in sum: "conscientious practice" [@sonning_replication_2021, p. 1182]

## The garden of forking paths {background-image="https://amor.cms.hu-berlin.de/~pallesid/teaching/forking_paths.jpeg" background-opacity=.5}

- or 'researcher degrees of freedom' [@simmons_false-positive_2011]
- the problem: there are many plausible ways to analyse any given data set
- there are many choices researchers make in:
  + experimental design
  + data collection
  + data preprocessing
  + data analyses
  + reporting
- the path we happen to go down can seem pre-determined [@gelman_statistical_nodate]
  + but can amount to HARKing, *p*-hacking, fishing
  
- the fastest solution: share everything and write transparently

# The current state of quantitative linguistics

- there is a trend towards empirical methods throughout linguistics
  + we should pay attention to methodological discussions in related fields
- we also find ourselves in a state of methdological crisis

## Kuhn's structure of scientific revoluations

- Thomas Kuhn's *The Theory of Scientifc Revoluations* (1962)
  + based on socio-historical observation
  + the evolution of scientificy theory is cycical
  + crisis leads to revolution
  
- also applies to research methodology
  
###

::: {.content-visible when-format="revealjs"}

:::: columns

::: {.column width="70%"}

Three recurrent phases:
  
1. **Normal science**
    + conventionally accepted set of procedures
    + researchers work within this paradigm
    + minor modifications to account for minor anomalies
2. **Crisis**
    + widespread recognition of larger challenges to current paradigm
    + questioning of conventionally accepted theory
3. **Revolution**
    + overthrowing of previous norms in favour of a new paradigm
    + leads to new normal science
:::

::: {.column width="30%"}
```{r eval = T, fig.env = "figure", out.width="100%", fig.align = "center", fig.pos="H", set.cap.width=T, fig.cap="Kuhn's cyclical evolution of scientific theory [Source: @sonning_replication_2021, Figure 1]"}
#| echo: false

library(magick)
magick::image_read(here::here("media/Soennig_Werner_2021_Kuhn_Figure1.png"))
```
:::
::::

:::

::: {.content-hidden when-format="revealjs"}
Three recurrent phases:
  
1. **normal science**
    + little controversy over theoretical underpinnings
    + researchers work on small problems within a theory
2. **crisis**
    + contradictions between theory and evidence
    + questioning of conventionally accepted theory
3. **revolution**
    + overthrowing of previous norms in favour of a new paradigm
    + leads to new normal science

```{r eval = T, fig.env = "figure", out.width="50%", fig.align = "center", fig.pos="H", set.cap.width=T, fig.cap="Download GitHub repositiory"}
#| echo: false

library(magick)
magick::image_read(here::here("media/Soennig_Werner_2021_Kuhn_Figure1.png"))
```
:::

## Previous cycles of statistical analyses

- proprietary, point-and-click software (e.g., SPSS)
  + move to open source programming languages (e.g., R, Python, Julia)

- ANOVAs
  + move to linear regression
  + then linear mixed models
    + random-intercepts only models
    + maximal models
    + parsimonious models
- now a trend towards Bayesian regression

## What do statisticians say?

- @wasserstein_moving_2019: list of Do's and Don't from statisticians
  + *Don't* base conclusions on *p*-values
  + *Do* think about ATOM: Accept uncertainty, be Thoughtful, Open, and Modest
- @wasserstein_asa_2016: the American Statistical Association's statement on *p*-values
  + *p*-values are often misused and misinterpreted
  + good statistical practice is part of good scientific practice
    + as such, relies on good study design and conduct
    + interpretation in context
    + complete and transparent reporting

# Overwhelmed? {background-image="https://amor.cms.hu-berlin.de/~pallesid/teaching/oprah_overwhelmed.jpeg" background-opacity=.5}

- we're in a state of crisis with a wealth of possible statistical paths
  + but no current "gold standard"
- this can lead to anxiety among researchers
  + which analysis should I run? am I doing it correctly?
- just keep in mind ATOM
  + strive for honesty, not perfection

# Revolution

:::: {.columns}

::: {.column width="60%"}
- methodological anxiety stems from shifting sands, but leads to revolution
- revolution usually comes from young newcomers
  + resistance to change usually comes those with more invested in the prior ways
- the good news: the revolution is underway
  + leads to an increase in resources and courses on e.g., multi-level models

- one suggested reform: Open Science!
:::

::: {.column width="40%"}

```{r}
#| echo: false
#| out-width: "95%"
magick::image_negate(magick::image_read(here::here("media", "revolution.jpeg")))
```

:::

::::

## The old vs. the new {.smaller}

:::: {.columns}

::: {.column width="50%"}

```{r}
#| echo: false
#| out-width: "95%"
#| fig-align: center
#| label: fig-soennig_table1
#| fig-cap: "Source: @sonning_replication_2021"
magick::image_negate(magick::image_read(here::here("media", "soennig_werner_2021_table1.png")))
```

:::

::: {.column width="50%"}

- changes refer to not only statistical analyses
  + but also emphasise transparency
- ideally, we (as a field) would up our analysis game
  + but a good first step is moving towards Open Science
  + share data, code
  + transparently map out your route in the 'garden of forking paths'
- these are steps we'll cover in this course
  + pre-registration
  + data and code sharing
  + reproducible workflow
  + transparent writing

:::

::::

# Simple fixes {background-image="https://amor.cms.hu-berlin.de/~pallesid/teaching/tools.jpeg" background-opacity=.5}

- planning and design
  + large sample sizes
  + establish pre-processing/analysis steps a priori
- methodologically
  + select variables based on theory and research questions
  + model non-independence of data points (mixed models)
  + move towards estimation and away from arbitrary significance thresholds
- writing
  + be transparent about choices made
  
## Words of comfort

> Less experienced scholars must not fear methodological attacks on their analyses, which are instead seen as informing interim interpretations that may require future modification.

--- @sonning_replication_2021, p. 1199

- in some sub-fields linear mixed models are still not considered the standard
  + so you're well situated despite the doom around *p*-values
- moving from frequentist (NHST) framework to the Bayesian framework is relatively painless
  + in this class we will run a LMM with `lme4` (Frequentist) and with `brms` (Bayesian)
  
# Running replications: what to replicate

:::: {.columns}

::: {.column width="50%"}
- what makes a study 'worth' replicating? 
  + suggestions from @isager_deciding_nodate:
    
  1. value/interest of the topic
  2. uncertainty about the claim
  3. quality of proposed replication
      + or ability to reduce uncertainty
  4. costs and feasibility

:::

::: {.column width="50%"}
- what makes a replication study 'worthy' for publication?
  + theoretical impact of the replicated finding
  + statistical power of the replication
:::
::::

## Replication value

- replication value (RV): "the expected utility of a finding before replication" [@isager_deciding_nodate, p. 6]
  + (scientific) value of the research claim
    + importance to the field, to policy, health etc.
  + the uncertainty of our knowledge about the claim
    + validity of study design, statistical power, bias, etc.

- replication aims to reduce uncertainty
  + which also increases utility of the claim
  
### Quantifying RV

- how to quantify *value* and *uncertainty*?

- @isager_replication_2021 suggest using...
  - average yearly citation count to estimate *value*
    + the more citations, the higher the impact the original study had
  - and sample size to estimate *uncertainty* ($\frac{1}{\sqrt{n}}$)
    + the higher the sample size, the more precise the estimate
    + the lower the sample size, the greater the uncertainty
    + i.e., $n$ is inversely correlated with *uncertainty*

## $RV_{Cn}$

:::: {.columns}

::: {.column width="60%"}

```{r}
#| echo: false
#| out-width: "95%"
#| fig-align: center
#| label: fig-RVcn
#| fig-cap: "Source: @isager_replication_2021, p. 25 (all rights reserved)"
magick::image_negate(magick::image_read(here::here("media/Isager_2021_RVcn.png")))
```

:::

::: {.column width="40%"}

@isager_replication_2021: 

$$
RV_{Cn} = value \ x \ uncertainty
$$

:::

::::

# Student replications {.smaller}

> It is peculiar that undergraduate students can be taught about the perils of underpowered studies in formal statistical instruction and simultaneously be required to perform research that is almost inevitably underpowered

--- @quintana_replication_2021, p. 1117

- a possible solution: student thesis replications
  + hands-on experience in open science practices
  + e.g., cumulative replication studies run by multiple groups
  
- some resources for students interested in replications
  + Student Theses Replication Network Linguistics [(STReNeL)](https://blog.junge-sprachwissenschaft.de/2023/01/16/strenel.html)
  + Collaborative REplications and Education Project [(CREP)](https://osf.io/wfc6u/)
  + Framework for Open and Reproducible Research Training [(FORRT)](https://forrt.org/)
  + German Reproducibility Network [(DERN)](https://reproducibilitynetwork.de/)

# Exercise

- Moodle: Quiz 'Kobrok & Roettger (2022)'
  + scan the article to answer the questions
  + this is not graded

# Learning objectives 🏁 {.unnumbered .unlisted .uncounted}

Today we learned...

- the replication crisis ✅
- replication in language sciences ✅
- requirements for replication ✅

# Important terms {.unnumbered .smaller .uncounted}


# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::


