---
title: "Data tidying"
subtitle: "From wide-to-long and long-to-wide with `tidyr`"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2024-05-14
format: 
  html:
    output-file: tidying.html
    number-sections: false
    toc: true
    code-overflow: wrap
    code-tools: true
    self-contained: true
  pdf:
    output-file: tidying.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  revealjs:
    output-file: tidying_slides.html
    include-in-header: ../../mathjax.html # for multiple equation hyperrefs
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    # chalkboard:
    #   src: chalkboard.json
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: false
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
    fig-align: center
    fig-dpi: 300
editor_options: 
  chunk_output_type: console
bibliography: ../../references.bib
csl: ../../apa.csl
execute:
  echo: true
---

# Learning objectives

Today we will...

- learn how to re-structure our data with the `tidyr` package
- use `pivot_longer()` to make data longer
- use `pivot_wider()` to make data wider

# Set-up {.unnumbered}

Load the `tidyverse` package

```{r}
library(tidyverse)
```

Load a subset of the `tidy_data_lifetime_pilot.csv` data. For demonstration purposes, we'll only look at two trials from a single participant.

```{r}
df_lifetime <- readr::read_csv(here::here("data/tidy_data_lifetime_pilot.csv"), 
                               # for special characters
                               locale = readr::locale(encoding = "latin1") 
                               ) |>
  filter(type=="critical", px=="px5", trial %in% c(3,8)) |> 
  select(px,trial,region,ff,fp,rpd,tt)
```

## Inspect data

- we'll be changing the shape of our data, so let's first see how it looks as-is

```{r}
df_lifetime 
```

- of importance, we have the following variables:
  + `region`: contains info on which sentence region the row's reading times correspond to
  + `ff`: first fixation time, an eye-tracking reading measure
  + `fp`: first-pass reading time, an eye-tracking reading measure
  + `rpd`: regression path duration, an eye-tracking reading measure
  + `tt`: total reading time, an eye-tracking reading measure
- we see that we have 10 rows x 4 reading time measures = 40 reading time measures


# Reshape data

- this is the major step of data tidying
  + make each column a variable
  + make each row an observation
  + make each cell a data point
- what variable and observation mean will depend on what you want to do, and will change at different steps of your analyses

- you typically want *long* data
  + but our dataset isn't as long as it could be

- the `tidyr` package from the `tidyverse` has some useful functions to faciliate this: `pivot_longer()` and `pivot_wider()`

## Pivot with `tidyr`

- to pivot (verb): *to turn or rotate on a point, like a hinge. Or a basketball player pivoting back and forth on one foot to protect the ball.* ([vocabulary.com](https://www.vocabulary.com/dictionary/pivot#:~:text=To%20pivot%20is%20to%20turn,foot%20to%20protect%20the%20ball.))

- a pivot (noun): *a fixed point supporting something that turns or balances* ([dictonary.Cambridge.org](https://dictionary.cambridge.org/dictionary/english/pivot))

::: {.content-visible when-format="revealjs"}
## 
:::

::: {.content-hidden when-format="pdf"}
```{r echo = F, fig.align = "center", out.width="100%"}
#| fig-cap-location: bottom
#| label: fig-friends
#| fig-cap: "A memorable scence (to millenials) from Friends where the word 'pivot' is repeatedly used ([YouTube clip](https://www.youtube.com/watch?v=8w3wmQAMoxQ))"

magick::image_read(here::here("media/pivot_friends.jpeg"))
```
:::

::: {.content-visible when-format="pdf"}
```{r echo = F, fig.align = "center", out.width="50%"}
#| fig-cap-location: bottom
magick::image_read(here::here("media/pivot_friends.jpeg"))
```
:::

# Wide-to-long: `pivot_longer()`

- `pivot_longer()` takes wide data and makes it longer
  + converts headers of columns into values of a new column
  + combines the values of those columns into a new condensed column
- takes a few arguments:
  + `cols`: which columns do we want to combine into a single column?
  + `names_to`: what should we call the new column containing the previous column names?
  + `values_to`: what should we call the new column containing the values from the previous columns?

::: {.content-visible when-format="revealjs"}
## `pivot_longer()`
:::

- let's take our four reading time measures and list them in a single variable that we'll call `measure`, and put their values in a second variable called `time`

```{r}
#| output-location: fragment
df_longer <-
  df_lifetime |> 
  pivot_longer(
  cols = c(ff,fp,rpd,tt), # columns to make long 
  names_to = "measure", # new column name for headers
  values_to = "time" # new column name for values
)
```

```{r}
df_longer
```


- now instead of having the four reading time values in a single row across four columns called `ff`, `fp`, `rpd`, and `tt`, we have two columns (`measure` and `time`) which contain the reading time measure names and corresponding reading times
- we again still have 40 reading time values: 40 rows x 1 column containing reading time values (`time`)

# Long-to-wide: `pivot_wider()`

- `pivot_wider()` takes long data and makes it wider
- takes a few arguments:
  + `id_cols`: identifying columns
  + `names_from`: what should we call the new column containing the previous column names?
  + `names_prefix`: 
  + `values_from`: new column values

::: {.content-visible when-format="revealjs"}
## `pivot_wider()`
:::

- let's now take our `region` column in `df_longer` and widen it
  + we'll do this only for `tt` (total reading time)
  the resultfour reading time measures and list them in a single variable that we'll call `measure`, and put their values in a second variable called `time`
  
```{r}
#| output-location: fragment
df_longer_wider <-
  df_longer |> 
  pivot_wider(
  id_cols = c(px,trial,measure), # columns to make long 
  names_from = region, # new column name for headers
  names_prefix = "reg_", # new column name for values (optional)
  values_from = time
) 
```

```{r}
df_longer_wider
```

- again, we have 40 reading time values: 8 rows x 5 variables containing reading time values per region

# Re-structuring, not changing

- in `df_lifetime`, `df_longer`, and `df_longer_wider`, we have 40 reading time values
  + we have the exact same information in all three versions
  + we have not removed or changed our data
  + we have only changed the *structure* of the data
- this might not always be the case, based on what you're trying to achieve
  + but it's important to understand that you can find the same information in long versus wide data
  + the way you structure your data should reflect/facilitate what you're trying to say about your data

::: {.content-visible when-format="revealjs"}
##
:::

- look at the three versions of the data below, and ask yourself: what does each one more easily communicate?

```{r}
df_longer
```

```{r}
# only first 15 rows
df_longer |> head(15)
```

```{r}
df_longer_wider
```



  
::: footer
More reading: [PsyTeachR](https://psyteachr.github.io/ads-v2/08-tidy.html#long-to-wide)
:::

# Learning objectives 🏁

Today we...

- learned how to re-structure our data with the `tidyr` package ✅
- used `pivot_longer()` to make data longer ✅
- used `pivot_wider()` to make data wider ✅