---
title: "Open Science Practices: Implementing a Reproducible Analysis Workflow"
subtitle: "SoSe2024"
author: "Daniela Palleschi"
institute: Humboldt-Universität zu Berlin
lang: en
date: 2023-04-12
format: 
  pdf:
    output-file: pdf-welcome.pdf
    toc: true
    number-sections: false
    colorlinks: true
    code-overflow: wrap
  html:
    output-file: sheet-welcome.html
    number-sections: false
    toc: true
    code-overflow: wrap
    code-tools: true
    self-contained: true
  revealjs:
    output-file: slides-welcome.html
    code-overflow: wrap
    theme: [dark]
    width: 1600
    height: 900
    progress: true
    scrollable: true
    # smaller: true
    slide-number: c/t
    code-link: true
    # logo: logos/hu_logo.png
    # css: logo.css
    incremental: true
    # number-sections: true
    toc: true
    toc-depth: 2
    toc-title: 'Overview'
    navigation-mode: linear
    controls-layout: bottom-right
    fig-cap-location: top
    font-size: 0.6em
    slide-level: 4
    self-contained: true
    title-slide-attributes: 
      data-background-image: logos/logos.tif
      data-background-size: 15%
      data-background-position: 50% 92%
editor_options: 
  chunk_output_type: console
bibliography: ../../references.bib
csl: ../../apa.csl
---

```{r setup, eval = T, echo = F}
knitr::opts_chunk$set(echo = F, # print chunks?
                      eval = T, # run chunks?
                      error = F, # print errors?
                      warning = F, # print warnings?
                      message = F, # print messages?
                      cache = F # cache?; be careful with this!
                      )
```

# Welcome!

- the language of instruction is English
- you will need a laptop in class from May 6h
- and a recent installation of 
    + R
    + RStudio
    + tinyTex
    
# Course Moodle

Course name: Open Science Practices: Implementing a Reproducible Analysis Workflow
Enrolment Key: r4repro

- course documents are available as HTML slides and PDF
- a course website will be up and running shortly

## Learning objectives

- learn basic concepts of Open Science in the scope of language research
  + focus on reproducibile analysis
- create and maintain self-contained analyses projects
- develop good habits for project management
- learn how to transparently write up analyses for a paper or thesis
- learn about concepts like version control and containerization

## What you will not learn in this book

- how to analyse data
- how to fit models
- how to plot data
- how to design an experiment

- we will not explicitly learn how to do these things, although we will be running simple code to achieve these goals
  
## Course credits

- 4 LP
  + 1LP: participation and readings
  + 1LP: in-class exercises and preparation
  + 1LP: 2 in-class quizzes (0.5LP each)
  + 1LP: a pre-registration (1LP)

# Session Info {.unlisted .unnumbered visibility="uncounted"}

Save your session info at the end of each document. Our results very often depend on the version of R/RStudio/a package we used. This is a great first step towards creating a reproducible workflow!

```{r}
sessionInfo()
```

# References {.unlisted .unnumbered visibility="uncounted"}

::: {#refs custom-style="Bibliography"}
:::
